Geometric and physical interpretation of the action principle

### 2023-06-27 Articles/2022-ActionPrinciple/ScientificReportsAmended/ActionPrincipleSR.tex
ActionPrinciple: Scientific Reports version

\section{Introduction}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will presents all the key point needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the appendix.

\section{One degree of freedom}

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that describes how states move in time.\footnote{Where possible, we will be writing the same expression in both vector calculus and component notations.} The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumptions requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free,\footnote{Given that this is a three dimensional space, we can use the standard tools of vector calculus.} that is
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning\footnote{In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.}
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative\footnote{We write the exterior derivative as $\partial \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.} of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions.

@@

@@

\section*{Appendix}

In this section we will review the arguments in more detail. Given that this article is still aimed at a physics audience, the focus will be conceptual rigor, rather than mathematical rigor. What is important is to establish a precise map between physical concepts and their mathematical representation, so that it is clear what physical requirement is captured by each mathematical definition.

We divide the appendix into three sections: the first will address the single d.o.f. case, the second will give a geometric interpretation of the tools of differential geometry that is suitable for physicists, and lastly we address the multiple d.o.f. case.

@@

This requirement tells us two things. First, that $X$ must be a differentiable manifold. To define densities over a set of variables and map them backward and forward in time Jacobians must be defined, which means the derivative across state variables must be defined as well, in particular with respect to time, which plays the double role of a coordinate of $X$ and the evolution parameter. This means that the evolutions must be differentiable curves and therefore the displacement $\vec{S}$ must be defined, which leads to
\begin{equation}
\tag*{(\ref{sdof_displacement})}
	\begin{aligned}
		\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
The double role of time gives us
\begin{equation}
\tag*{(\ref{sdof_time_constraint})}
 	S^t=\frac{dt}{dt}=1.
 \end{equation}

@@

@@

Now that we have developed a clear geometric and physical understanding of Hamilton's equations, we can concentrate on the action principle. For (iv) we can apply Stokes' theorem. We take a path $\gamma$ and a variation $\gamma'$. Since they share the endpoints, taken together they form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\vec{S}$ is divergence-free, the flow of $\vec{S}$ through a surface depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we have
\begin{equation}
\tag*{(\ref{sdof_action})}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &=
	\iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&= \oint_{\partial \Sigma} \vec{\theta}  \cdot d\vec{\gamma} \\
	&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

Now, since every functional has a corresponding form, $\partial F$ must have a corresponding $k+1$-form which must be fully identified by $\omega$. We note it as $\partial \omega$ and write
\begin{equation*}
	\partial F(\Sigma) = \int_\Sigma \partial \omega(d\Sigma) = \oint_{\sigma = \partial \Sigma} \omega(d\sigma) = F(\partial \Sigma).
\end{equation*}
This is the generalized Stokes' theorem and $\partial \omega$ is the exterior derivative of $\omega$. We use $\partial$ as we leave $d$ reserved for the infinitesimal regions.

@@

Moving on, the identities
\begin{equation*}
	\begin{aligned}
		\nabla \cdot \nabla \times \vec{f} &= 0 \\
		\nabla \times \nabla \varphi &= 0
	\end{aligned}
\end{equation*}
both correspond to the single identity
\begin{equation*}
	\partial \partial \omega = 0,
\end{equation*}
which states that the exterior derivative applied twice is also zero. This identity can be understood in terms of the functionals. We have  $\partial \partial F(\Sigma) = F(\partial \partial \Sigma)$ and the boundary of a boundary $\partial \partial \Sigma = \emptyset$ is always the empty set. So, no matter what $F$ or $\Sigma$ is, we are evaluating a functional on an empty set, which is zero.

In the same way that a divergence-free field can be written as the curl of a vector potential, or a curl-free field can be written as the gradient of a scalar potential, a form whose exterior derivative is zero $\partial \omega = 0$ (a closed form), can be written as the exterior derivative $\omega = \partial \theta$ of a lower dimensional form, at least in some neighborhood. In terms of functionals, this corresponds to the case where $F(\sigma)=0$ for all contractible surfaces, those surfaces that can be morphed to a point with a continuous transformation. If $F(\sigma)=0$ over all surfaces, instead, then the corresponding form is exact.

@@

As in the previous case, assumption \ref{assum_detrev} tells us that the integral over a closed hyper-surface is zero, since as many states flow in as flow out. We have
\begin{equation}
	\begin{aligned}
		\oint_{\Sigma = \partial V} \Omega(d\Sigma) &= 0 = \int_V \partial \Omega(dV) \\
		\partial\Omega &= 0.
	\end{aligned}
\end{equation}
which tells us that $\Omega$ is closed.

@@

Using assumptions \ref{assum_detrev} and \ref{assum_indep} again we can see how $\omega$ must be closed
\begin{equation}
	\partial \omega = 0.
\end{equation}
If we imagine translating a surface along an independent variable, the number of configurations identified by the surface cannot change. This uses \ref{assum_indep}. If we imagine translating in time, again the number of configurations cannot change because of \ref{assum_detrev}. Therefore, if we imagine a parallelepiped, opposite sides must identify the same number of configurations. The integral will take the sides with an opposite orientation, and therefore there will be zero net contribution. It is interesting to note that this condition requires both assumptions.

Putting it all together, assuming \ref{assum_detrev} and \ref{assum_indep} gives us the existence of the displacement vector field $\vec{S}$ and the closed counting two-form $\omega$ such that $\vec{S}$ identifies the only vector field that kills $\omega$.

The converse is also true: if we start with the mathematical requirements, we can recover the physical assumptions. If we are given a two-form $\omega$ on a $2N+1$ dimensional space, we can define the $2N$ form $\Omega = \omega^N$. This form will be non-zero because $\vec{S}$ is the only vector field that kills the two-form $\omega$. It will also be closed since $\partial \omega = 0$ and
\begin{equation}
	\partial \Omega = N\partial \omega \wedge \omega^{N-1} = 0.
\end{equation}
So the flow of the well defined displacement field $\vec{S}$, as measured by $\Omega$, is zero through closed surfaces. This implies \ref{assum_detrev}.

@@

For (ii), we note that since the form is closed, we can find, at least in a contractible region, a covector potential $\theta$ such that $\omega=\partial \theta$. Those familiar with symplectic and contact geometry will know that using Darboux's theorem, we can find a set of coordinates such that we can express $\theta$ as
\begin{equation}
	\tag*{(\ref{mdof_potential_expression})}
	\theta = p_i e^{q^i} - H e^t.
\end{equation}
Just stating this is, of course, not very insightful. To gain more insight, we provide the sketch of a proof that, as we have done for the single degree of freedom, uses gauge transformations to fix the form of the potential.

As a starting point, we take conditions \ref{canonical_conditions} that state mathematically that $q^i$ and $p_i$ form independent degrees of freedom. We re-express those conditions in terms of the potential $\theta$
\begin{equation}\label{canonical_potential_conditions}
\begin{aligned}
	\omega(e^{q^i}, e^{p_j}) &= (-\partial\theta)_{q^i p_j} = -(\partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i}) = \delta^i_j \\
	\omega(e^{q^i}, e^{q^j}) &= (-\partial\theta)_{q^i q^j} = -(\partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i}) = 0 \\
	\omega(e^{p_i}, e^{p_j}) &= (-\partial\theta)_{p_i p_j} = -(\partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i}) = 0
\end{aligned}
\end{equation}
We can use our gauge freedom to set $\theta_{p_1} = 0$, much in the same way we did for the simpler case. We now have $\partial_{q^1} \theta_{p_1} = 0$ and, by the first condition, $\partial_{p_1} \theta_{q^1} = 1$. Integrating, we have $\theta_{q^1} = p_1 + g(q^i, p_2, p_3, ..., t)$ where $g$ is an arbitrary function which we can set to zero. Like in the single degree of freedom, we can do this because $g$ does not depend on $p_1$, so it can be eliminated with a gauge transformation that does not change $\theta_{p_1}$. Therefore we have:
\begin{equation}
	\theta = p_1 e^{q^1} + 0 e^{p_1} + \theta_{q^2} e^{q^2} + \theta_{p_2} e^{p_2} + ... + \theta_{t} e^{t}.
\end{equation}

@@

The components of $\partial\theta$ are
\begin{equation}
	\begin{aligned}
		(\partial\theta)_{q^i p_j} &= \partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i} = \partial_{q^i} 0 - \partial_{p_j} p_i= - \delta^i_j \\
		(\partial\theta)_{q^i q^j} &= \partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i} = \partial_{q^i}p_j - \partial_{q^j}p_i = 0 \\
		(\partial\theta)_{p_i p_j} &= \partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i} = \partial_{p_i} 0 - \partial_{p_j} 0 = 0 \\
		(\partial\theta)_{q^i t} &= \partial_{q^i}\theta_{t} - \partial_{t}\theta_{q^i} = \partial_{q^i} (-H) - \partial_{t} p_i= - \partial_{q^i} H \\
		(\partial\theta)_{p_i t} &= \partial_{p_i}\theta_{t} - \partial_{t}\theta_{p_i} = \partial_{p_i} (-H) - \partial_{t} 0= - \partial_{p_i} H \\
	\end{aligned}
\end{equation}
Therefore we recover
\begin{equation}
	\tag*{(\ref{mdof_form_components})}
	\omega_{ab}=(-\partial\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		- \partial_{q^j} H & - \partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

For a single degree of freedom, recalling that $S^t = 1$, we can describe $\omega$ with the following nine components
\begin{equation}\label{mdof_reduceto_sdof}
	\omega_{ab}=(-\partial\theta)_{ab} = \begin{bmatrix}
		0 & S^t & -S^p \\
		-S^t & 0 & S^q \\
		S^p & - S^q & 0
	\end{bmatrix}.
\end{equation}
In three dimensions, using the Levi-Civita symbol $\epsilon^{abc}$ we can write
\begin{equation}
	S^a = \epsilon^{abc} \omega_{bc}
\end{equation}
which tells us that $\vec{S}$ is the pseudo-vector corresponding to the direction perpendicular to $\omega$. This is the same relationship that we would have, for example, for the magnetic field psuedo-vector and the spatial components of the electromagnetic field in relativity. If we express $\omega$ in terms of $\vec{\theta}$, we have
\begin{equation}
	S^a = - \epsilon^{abc} \partial_b \theta_c
\end{equation}
which recovers \ref{sdof_displacement_potential}: the displacement field is minus the curl of $\omega$. Thus we can see exactly how the general case reduces to the single d.o.f. case.

For (iv), we simply need to apply Stokes' theorem. The situation is identical to the single degree of freedom, except that the flow now is not given by the surface integral of $\vec{S}$, but of the form $\omega$. The path $\gamma$ and its variation $\gamma'$ will form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\omega$ is closed, its surface integral depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Putting it all together we recover
\begin{equation}
\tag*{(\ref{mdof_action})}
\begin{aligned}
	- \int_{\Sigma} \omega(d\Sigma) &=
	\int_{\Sigma} \partial \theta (d\Sigma) \\
	&= \oint_{\partial \Sigma = \gamma \cup \gamma'} \theta(d\gamma) \\
	&= \int_{\gamma} \theta(d\gamma) - \int_{\gamma'} \theta (d\gamma') \\
	&= \delta \int_{\gamma} \theta (d\gamma).
\end{aligned}
\end{equation}

>>>>

\section*{Introduction}

@@

The mathematics needed to run the argument is well established\cite{souriau1970structure, abraham1978foundations,arnold1989mathematical, marsden1999introduction}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometric interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will present all the key point needed to follow the argument and its physical and geometric meaning, leaving the mathematical details and calculations to the Supplementary Information.

\section*{One degree of freedom}

@@

Under the assumption that
\begin{align}\label{assum_detrev}
	\tag*{(DR)}
	\parbox{2.8in}{\emph{the system undergoes deterministic and reversible evolution}}
\end{align}
we can define a displacement vector field
\begin{equation}\label{sdof_displacement}
\begin{aligned}
	\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
	&= S^a e_a = \frac{d\xi^a}{dt} e_a .
\end{aligned}
\end{equation}
that describes how states move in time. [Where possible, we will be writing the same expression in both vector calculus and component notations.] The time component of the displacement vector field is constrained, as we have
\begin{equation}\label{sdof_time_constraint}
	S^t=\frac{dt}{dt}=1.
\end{equation}

If assumption \ref{assum_detrev} is valid, we expect the flow of states through a closed surface to be zero: as many states flow in as flow out of the region. Alternatively, if we assign a probability, or probability density, to each trajectory, the assumptions requires that probability not to change, so integrating the probability over a closed surface must yield zero. However we see it, assumption \ref{assum_detrev} means the field is divergence-free.  [Given that this is a three dimensional space, we can use the standard tools of vector calculus.] That is, 
\begin{equation}\label{sdof_div_free}
	\nabla \cdot \vec{S} = \partial_a S^a = 0.
\end{equation}

@@

The form $\omega$ will need to be anti-symmetric
\begin{equation}\label{mdof_form_antisymm}
	\omega(\vec{v}, \vec{w}) = - \omega(\vec{w}, \vec{v})
\end{equation}
as the parallelogram identified by $\vec{v}$ and $\vec{w}$ in that order will be the same as the one identified by $\vec{w}$ and $\vec{v}$ with opposite orientation. The form $\omega$ will also be closed, meaning 
\begin{equation}\label{mdof_closed_form}
	\oiint_\Sigma \omega(d\Sigma) = 0
\end{equation}
over all closed contractible surfaces. [In our notation, $d\Sigma$ represents the infinitesimal surface element of integration, which is the argument of the form.] This stems from \ref{assum_indep}. Suppose you take a surface and translate it along another independent variable. The count of states cannot depend on the independent variable. Therefore if we construct any parallelepiped at equal time, two opposite sides will contain the same number of states with opposite orientation, and therefore the integral of $\omega$ over the whole surface will be zero.

Since $\omega$ is closed, in every contractible region it can be expressed as the exterior derivative of a covector $\theta$:
\begin{equation}\label{mdof_form_potential}
	\begin{aligned}
		\theta &= \theta_a e^a = \theta_{q^i} e^{q^i} + \theta_{p_i} e^{p_i} + \theta_t e^t \\
		\omega &= - \partial \theta = - \left( \partial_a \theta_b - \partial_b \theta_a \right) e^a \otimes e^b
	\end{aligned}
\end{equation}
The minus sign is required to match conventions. [We write the exterior derivative as $\partial \omega$ instead of $d \omega$ so that $d$ is only used for infinitesimal line or surface elements.]

@@

\vspace{-1mm}
\section*{Data availability}
Not applicable.

@@

\newpage
\section*{Supplementary Information}
\setcounter{page}{1}

In this Supplementary Information we will review the arguments in more detail. Given that this article is still aimed at a physics audience, the focus will be conceptual rigor, rather than mathematical rigor. What is important is to establish a precise map between physical concepts and their mathematical representation, so that it is clear what physical requirement is captured by each mathematical definition.

We divide the material into three sections: the first will address the single d.o.f. case, the second will give a geometric interpretation of the tools of differential topology that is suitable for physicists, and lastly we address the multiple d.o.f. case.

@@

This requirement tells us two things. First, that $X$ must be a differentiable manifold. To define densities over a set of variables and map them backward and forward in time Jacobians must be defined, which means the derivative across state variables must be defined as well, in particular with respect to time, which plays the double role of a coordinate of $X$ and the evolution parameter. This means that the evolutions must be differentiable curves and therefore the displacement $\vec{S}$ must be defined, which leads to
\begin{equation}
\tag*{(\ref{sdof_displacement})}
	\begin{aligned}
		\vec{S} &= \left[ \frac{dq}{dt},\frac{dp}{dt},\frac{dt}{dt} \right] \\
		&= S^a e_a = \frac{d\xi^a}{dt} e_a .
	\end{aligned}
\end{equation}
In the literature of dynamical systems, the displacement vector field $\vec{S}$ is simply called the vector field of the dynamical system. The expression of the vector field is what determines the equations of motion in a dynamical system. The double role of time gives us
\begin{equation}
\tag*{(\ref{sdof_time_constraint})}
 	S^t=\frac{dt}{dt}=1.
 \end{equation}

@@

The potential $\vec{\theta}$ is closely related to the canonical one-form of symplectic geometry and the contact form of contact geometry, but there are some differences. Both the canonical one-form and the contact form are a specific covector, defined with specific coordinates. For us, $\vec{\theta}$ is a potential and therefore is not fully specified by the physics. Writing down a specific potential means choosing a gauge, therefore our potential corresponds to the canonical one-form or the contact form only if a specific gauge is fixed.

@@

Now that we have developed a clear geometric and physical understanding of Hamilton's equations, we can concentrate on the action principle. For (iv) we can apply Stokes' theorem. We take a path $\gamma$ and a variation $\gamma'$. Since they share the endpoints, taken together they form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\vec{S}$ is divergence-free, the flow of $\vec{S}$ through a surface depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Thus we have
\begin{equation}
\tag*{(\ref{sdof_action})}
\begin{aligned}
	- \iint_{\Sigma} \vec{S} \cdot d\vec{\Sigma} &=
	\iint_{\Sigma} \left( \nabla \times \vec{\theta} \right) \cdot d\vec{\Sigma} \\
	&= \oint_{\partial \Sigma} \vec{\theta}  \cdot d\vec{\gamma} \\
	&= \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma} - \int_{\gamma'} \vec{\theta} \cdot d\vec{\gamma}' \\
	&= \delta \int_{\gamma} \vec{\theta} \cdot d\vec{\gamma}.
\end{aligned}
\end{equation}

@@

Now, since every functional has a corresponding form, $\partial F$ must have a corresponding $k+1$-form which must be fully identified by $\omega$. We note it as $\partial \wedge \omega$ and write
\begin{equation*}
	\partial F(\Sigma) = \int_\Sigma \partial \wedge \omega(d\Sigma) = \oint_{\sigma = \partial \Sigma} \omega(d\sigma) = F(\partial \Sigma).
\end{equation*}
This is the generalized Stokes' theorem and $\partial \wedge \omega$ is the exterior derivative of $\omega$. We use $\partial$ as we leave $d$ reserved for the infinitesimal regions.

@@

Moving on, the identities
\begin{equation*}
	\begin{aligned}
		\nabla \cdot \nabla \times \vec{f} &= 0 \\
		\nabla \times \nabla \varphi &= 0
	\end{aligned}
\end{equation*}
both correspond to the single identity
\begin{equation*}
	\partial \wedge \partial \wedge \omega = 0,
\end{equation*}
which states that the exterior derivative applied twice is also zero. This identity can be understood in terms of the functionals. We have  $\partial \partial F(\Sigma) = F(\partial \partial \Sigma)$ and the boundary of a boundary $\partial \partial \Sigma = \emptyset$ is always the empty set. So, no matter what $F$ or $\Sigma$ is, we are evaluating a functional on an empty set, which is zero.

In the same way that a divergence-free field can be written as the curl of a vector potential, or a curl-free field can be written as the gradient of a scalar potential, a form whose exterior derivative is zero $\partial \wedge \omega = 0$ (a closed form), can be written as the exterior derivative $\omega = \partial \wedge \theta$ of a lower dimensional form, at least in some neighborhood. In terms of functionals, this corresponds to the case where $F(\sigma)=0$ for all contractible surfaces, those surfaces that can be morphed to a point with a continuous transformation. If $F(\sigma)=0$ over all surfaces, instead, then the corresponding form is exact.

@@

As in the previous case, assumption \ref{assum_detrev} tells us that the integral over a closed hyper-surface is zero, since as many states flow in as flow out. We have
\begin{equation}
	\begin{aligned}
		\oint_{\Sigma = \partial V} \Omega(d\Sigma) &= 0 = \int_V \partial \wedge \Omega(dV) \\
		\partial \wedge\Omega &= 0.
	\end{aligned}
\end{equation}
which tells us that $\Omega$ is closed.

@@

Using assumptions \ref{assum_detrev} and \ref{assum_indep} again we can see how $\omega$ must be closed
\begin{equation}
	\partial \wedge \omega = 0.
\end{equation}
If we imagine translating a surface along an independent variable, the number of configurations identified by the surface cannot change. This uses \ref{assum_indep}. If we imagine translating in time, again the number of configurations cannot change because of \ref{assum_detrev}. Therefore, if we imagine a parallelepiped, opposite sides must identify the same number of configurations. The integral will take the sides with an opposite orientation, and therefore there will be zero net contribution. It is interesting to note that this condition requires both assumptions.

The form $\omega$ corresponds to the symplectic form of symplectic manifolds and of the symplectic bundles of contact manifolds with the following caveats. As we are in phase space extended by time, we have a contact manifold, and $\omega$ mathematically corresponds to the exterior derivative of the contact form, which is the primitive mathematical object. Physically, this is backwards: $\omega$ is the actual primitive physical object and the contact form $\theta$ is a potential, which is unphysical as it is fixed by an arbitrary gauge. Note how the mathematical definitions make it hard to have a coherent physical interpretation. The primitive objects for contact manifolds and symplectic manifolds are different, $\theta$ in the first case and $\omega$ in the other, and they are defined and studied as two separate mathematical spaces, while physically they are the same space with or without time.

To summarize, assuming \ref{assum_detrev} and \ref{assum_indep} gives us the existence of the displacement vector field $\vec{S}$ and the closed counting two-form $\omega$ such that $\vec{S}$ identifies the only vector field that kills $\omega$.

The converse is also true: if we start with the mathematical requirements, we can recover the physical assumptions. If we are given a two-form $\omega$ on a $2N+1$ dimensional space, we can define the $2N$ form $\Omega = \omega^N$. This form will be non-zero because $\vec{S}$ is the only vector field that kills the two-form $\omega$. It will also be closed since $\partial \wedge \omega = 0$ and
\begin{equation}
	\partial \wedge \Omega = N\partial \wedge \omega \wedge \omega^{N-1} = 0.
\end{equation}
So the flow of the well defined displacement field $\vec{S}$, as measured by $\Omega$, is zero through closed surfaces. This implies \ref{assum_detrev}.

@@

For (ii), we note that since the form is closed, we can find, at least in a contractible region, a covector potential $\theta$ such that $\omega=\partial \wedge \theta$. Those familiar with symplectic and contact geometry will know that using Darboux's theorem, we can find a set of coordinates such that we can express $\theta$ as
\begin{equation}
	\tag*{(\ref{mdof_potential_expression})}
	\theta = p_i e^{q^i} - H e^t.
\end{equation}
Just stating this is, of course, not very insightful. To gain more insight, we provide the sketch of a proof that, as we have done for the single degree of freedom, uses gauge transformations to fix the form of the potential.

As a starting point, we take conditions \ref{canonical_conditions} that state mathematically that $q^i$ and $p_i$ form independent degrees of freedom. We re-express those conditions in terms of the potential $\theta$
\begin{equation}\label{canonical_potential_conditions}
\begin{aligned}
	\omega(e^{q^i}, e^{p_j}) &= (-\partial \wedge\theta)_{q^i p_j} = -(\partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i}) = \delta^i_j \\
	\omega(e^{q^i}, e^{q^j}) &= (-\partial \wedge\theta)_{q^i q^j} = -(\partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i}) = 0 \\
	\omega(e^{p_i}, e^{p_j}) &= (-\partial \wedge\theta)_{p_i p_j} = -(\partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i}) = 0
\end{aligned}
\end{equation}
We can use our gauge freedom to set $\theta_{p_1} = 0$, much in the same way we did for the simpler case. We now have $\partial_{q^1} \theta_{p_1} = 0$ and, by the first condition, $\partial_{p_1} \theta_{q^1} = 1$. Integrating, we have $\theta_{q^1} = p_1 + g(q^i, p_2, p_3, ..., t)$ where $g$ is an arbitrary function which we can set to zero. Like in the single degree of freedom, we can do this because $g$ does not depend on $p_1$, so it can be eliminated with a gauge transformation that does not change $\theta_{p_1}$. Therefore we have:
\begin{equation}
	\theta = p_1 e^{q^1} + 0 e^{p_1} + \theta_{q^2} e^{q^2} + \theta_{p_2} e^{p_2} + ... + \theta_{t} e^{t}.
\end{equation}

@@

The components of $\partial \wedge\theta$ are
\begin{equation}
	\begin{aligned}
		(\partial \wedge\theta)_{q^i p_j} &= \partial_{q^i}\theta_{p_j} - \partial_{p_j}\theta_{q^i} = \partial_{q^i} 0 - \partial_{p_j} p_i= - \delta^i_j \\
		(\partial \wedge\theta)_{q^i q^j} &= \partial_{q^i}\theta_{q^j} - \partial_{q^j}\theta_{q^i} = \partial_{q^i}p_j - \partial_{q^j}p_i = 0 \\
		(\partial \wedge\theta)_{p_i p_j} &= \partial_{p_i}\theta_{p_j} - \partial_{p_j}\theta_{p_i} = \partial_{p_i} 0 - \partial_{p_j} 0 = 0 \\
		(\partial \wedge\theta)_{q^i t} &= \partial_{q^i}\theta_{t} - \partial_{t}\theta_{q^i} = \partial_{q^i} (-H) - \partial_{t} p_i= - \partial_{q^i} H \\
		(\partial \wedge\theta)_{p_i t} &= \partial_{p_i}\theta_{t} - \partial_{t}\theta_{p_i} = \partial_{p_i} (-H) - \partial_{t} 0= - \partial_{p_i} H \\
	\end{aligned}
\end{equation}
Therefore we recover
\begin{equation}
	\tag*{(\ref{mdof_form_components})}
	\omega_{ab}=(-\partial \wedge\theta)_{ab} = \begin{bmatrix}
		0 & \delta^j_i & \partial_{q^i} H \\
		-\delta^i_j & 0 & \partial_{p_i} H \\
		- \partial_{q^j} H & - \partial_{p_j} H & 0
	\end{bmatrix}.
\end{equation}

@@

For a single degree of freedom, recalling that $S^t = 1$, we can describe $\omega$ with the following nine components
\begin{equation}\label{mdof_reduceto_sdof}
	\omega_{ab}=(-\partial \wedge\theta)_{ab} = \begin{bmatrix}
		0 & S^t & -S^p \\
		-S^t & 0 & S^q \\
		S^p & - S^q & 0
	\end{bmatrix}.
\end{equation}
In three dimensions, using the Levi-Civita symbol $\epsilon^{abc}$ we can write
\begin{equation}
	S^a = \epsilon^{abc} \omega_{bc}
\end{equation}
which tells us that $\vec{S}$ is the pseudo-vector corresponding to the direction perpendicular to $\omega$. This is the same relationship that we would have, for example, for the magnetic field psuedo-vector and the spatial components of the electromagnetic field in relativity. If we express $\omega$ in terms of $\vec{\theta}$, we have
\begin{equation}
	S^a = - \epsilon^{abc} \partial_b \theta_c
\end{equation}
which recovers \ref{sdof_displacement_potential}: the displacement field is minus the curl of $\omega$. Thus we can see exactly how the general case reduces to the single d.o.f. case.

For (iv), we simply need to apply Stokes' theorem. The situation is identical to the single degree of freedom, except that the flow now is not given by the surface integral of $\vec{S}$, but of the form $\omega$. The path $\gamma$ and its variation $\gamma'$ will form a closed curve where $\gamma'$ is taken with the opposite orientation. Because $\omega$ is closed, its surface integral depends only on its contour. Moreover, by Stokes' theorem, the flow through the surface will be equal to the integral of the potential over the contour. Thus we have
\begin{equation}
\tag*{(\ref{mdof_action})}
\begin{aligned}
	- \int_{\Sigma} \omega(d\Sigma) &=
	\int_{\Sigma} \partial \wedge \theta (d\Sigma) \\
	&= \oint_{\partial \Sigma = \gamma \cup \gamma'} \theta(d\gamma) \\
	&= \int_{\gamma} \theta(d\gamma) - \int_{\gamma'} \theta (d\gamma') \\
	&= \delta \int_{\gamma} \theta (d\gamma).
\end{aligned}
\end{equation}
